Normalizations are not what you think; Explicit Scale Simulation in ALDEx2

### 2024-08-13 go3.Rmd
final go3 before Justin

The analysis of  HTS data suffers from several known problems that can be
traced, in whole or in part, to  misspecification of scale. The first issue 
is poor control of the false discovery rate (FDR) [@Thorsen:2016aa;@Quinn:2018aa], 
  exhibited as dataset-dependent 
FDR control.  The FDR problem is connected to the double-filtering approach
that is often used, but which is known not to have appropriate FDR control 
[@Zhang:2009aa;@Ebrahimpoor:2021aa]. The second issue is poor performance when 
analyzing asymmetric data; data that arises because of a bias in the direction of change.
This type of data frequently arises in in-vitro selection experiments (SELEX), 
transcriptome analysis, and microbiome analysis [@Wu2021]. The third issue is that these
problems become more pronounced as more samples are collected; that is,
more information results in a worsening of the accuracy of the analysis [@nixon2023scale,@Nixon2024B]. 
The final problem is that these data are relative data, i.e. compositional, 
and substantial effort has been put into removing this constraint.

@@

Recently, Nixon et al. [-@nixon2023scale] showed that the challenge of non-biological variation in sequencing depth be viewed as a problem of partially-identified models. They showed that \emph{all} normalizations make some assumption about  scale. These implicit assumptions are often  difficult to interpret, and different normalizations provide different outputs when applied to the same dataset [@Bullard:2010;@Dillies:2013;@Thorsen:2016aa;@Weiss:2017aa]. Intuitively, normalizations in widespread use  assume that either all samples have the same scale, e.g. proportions, rarefaction [@Hughes:2005tu], RPKM [@Mortazavi:2008; @wagner:tpm], etc; or that a subset of features in one sample can be chosen as a reference to which the others are scaled e.g. the TMM,[@Robinson:2010a], or LVHA [@Wu2021]; or that different sub-parts of each sample maintain a constant scale across samples e.g. the RLE [@Anders:2010];  or that the geometric mean of the parts is appropriate e.g the CLR [@aitchison1982] and its derivatives.

The original ALDEx2 [@fernandes:2013] model made a strict assumption about scale through the CLR normalization [@nixon2023scale]. While this assumption could be useful in many cases it could  never be exactly true, and others have shown that it is not always the best option [@Yerke:2024aa]. Nixon et al. [-@nixon2023scale] showed that better scale assumptions resulted in more reproducible data analysis including better control of both false positive and false negative results. In essence ALDEx2 has been modified to explicitly model the scale over a range of reasonable normalization parameters. Here, we briefly introduce these modifications and then show how scale uncertainty can greatly improve modeling in transcriptome and meta-transcriptome datasets to provide more robust and reproducible results.

@@

All  the normalizations in current use are data transformations that can be stated as ratios of the form \(\hat{{\mathbf{W}}}_{dn}=\mathbf{Y}_{dn}/f(\mathbf{Y})\), where the denominator is determined by some function of the observation and the \(\hat{{\mathbf{W}}}\) notation indicates that the normalization is attempting to provide an estimate of the true value.  The technical variation in sequencing depth (\(\mathbf{Y}^{\perp}_{n}=\sum_{d}\mathbf{Y}_{dn}\)) implies that observed data \(\mathbf{Y}\) provides us with information about the system composition \(\mathbf{W}^{\parallel}\) but little to no information in the system scale \(\mathbf{W}^{\perp}\) (Lovell et al. 2011).

@@

The ALDEx2 R package [@fernandes:2013] is a general purpose toolbox for Bayesian modeling of HTS data. For brevity, we discuss ALDEx2 in its simplest form as a tool for estimating the magnitude and statistical significance of Log-Fold-Changes (LFC) (e.g., differential abundance or differential expression analysis), but note that it can be used to fit more complex linear models. At a high-level, ALDEx2 involves three steps: 1) a Bayesian model is used to estimate \(\mathbf{\hat{W}}^{\parallel}\) given observations \(\mathbf{Y}\); 2) a normalization is used to estimate \(\mathbf{\hat{W}}\) given the estimate \(\mathbf{\hat{W}}^{\parallel}\); 3) the \(\mathbf{\hat{W}}\) estimates are used to estimate LFC as part of differential abundance (or expression) analyses. For more details on ALDEx2 see [@fernandes:2013;@nixon2023scale].

By default, ALDEx2 uses the CLR normalization which can be written as \(\log \mathbf{\hat{W}}_{dn}=\log (\mathbf{\hat{W}}^{\parallel}_{dn}/G_{n})\) where \(G_{n}\) denotes the geometric mean of the vector \((\hat{W}^{\parallel}_{1n}, \dots, \hat{W}^{\parallel}_{Dn})\). The CLR normalization equates to an implicit assumption that \(\mathbf{\hat{W}}^{\perp}_{n}=1/G_{n}\)[@nixon2023scale,Nixon2024B] and even slight errors in this assumption introduces bias into the LFC estimate that is not accounted for when estimating uncertainty (e.g., confidence intervals or credible sets). In fact, Nixon et al. [-@nixon2023scale] showed that the only way in which the ALDEx2 model, or any normalization-based model, could ever be calibrated (e.g., control Type-I Error rates) was if this assumption was exactly true.  In support of this assertion it is known that false discovery rates vary widely by analysis, dataset, and normalization method [@hawinkel2017;@Li:2022aa,Nixon2024B]. To fix this, Nixon et al. [-@nixon2023scale] showed that models should incorporate potential error in the assumptions implied by normalizations.

@@

Many groups have conducted benchmarking studies on different tools and normalizations used for the analysis of datasets such as transcriptomes[@Bullard:2010;@Soneson:2013;@Schurch:2016aa;@Quinn:2018aa] and microbiomes [@McMurdie:2014a;@Thorsen:2016aa; @Weiss:2017aa; @hawinkel2017;@Yerke:2024aa]. Generically, it is observed that the actual agreement between analysis methods can be modest, and  no single method outperforms all others in every dataset or type of experiment.  That different tools appear to work more reliably in different datasets from different sources can be explained as different  normalizations being a better fit to the scale  a particular dataset by chance.

From our perspective, the disagreement between tools can be explained by the partially identifiable nature of the data; that is,  the  observed data are consistent with multiple ways of estimating the parameters [@nixon2023scale]. Partially identifiable data exhibit multiple pathologies, chief among them being that that different analytic approaches can produce  different parameter estimates and that more data produces worse estimates because the additional data increases the precision of a flawed estimate [@gustafson2015bayesian].  In the analysis of HTS data it is often observed that larger datasets converge on the majority of parts being significantly different [@Schurch:2016aa;@nixon2023scale], and that different analytic approaches result in different parts being chosen as significantly different [@Soneson:2013;@Schurch:2016aa;@Quinn:2018aa].

Scale simulation  is now build into ALDEx2 to address the issue of partially identifiability, and through this lens the two main root causes to common HTS data pathologies can be proposed. The first contributing factor is the observed very low dispersion estimate for many features that is a by-product of experimental design, sequencing and normalization. In the Schurch et al. @Schurch:2016aa dataset, the data were  from single colonies derived from a single culture. Thus, it is more accurate to describe the 96 samples as wet-lab technical replicates rather than independent samples. However, this type of replicate is standard in the molecular literature, and would be expected to result in the very low dispersion that is observed.

@@

While the actual scale of the underlying environment is inaccessible post-sequencing, we can study the sensitivity of the results to the choice of normalization and thus scale.
To do this we incorporate scale models [@nixon2023scale] in lieu of normalizations through the default scale model that is now built into ALDEx2.
Scale uncertainty is incorporated using the \texttt{gamma} parameter that controls the amount uncertainty added to the CLR mean assumption when we call either \texttt{aldex()}, or \texttt{aldex.clr()}.
The ALDEx2 package  contains a sensitivity analysis function, \texttt{aldex.senAnalysis()}, that can be used to explore the effect of different amounts of scale uncertainty, and an example is shown in  Supplementary Figure 2 for the yeast transcriptome dataset. Here it is clear that even tiny amounts of scale preclude the majority of transcripts from being considered statistically significant.
In practice, we suggest that a \texttt{gamma} parameter between 0.5 and 1 is realistic for most experimental designs, but further work is needed.

@@

Beyond concerns of fidelity and rigor, scale models also enhance the reproducibility and transparency of HTS analyses.
An advantage of incorporating scale is that analyses can be made much more robust such that actual or potential differences in scale can be tested and accounted for explicitly.
This occurs because rather than using an implicit assumption (e.g., \(\log \mathbf{W}^{\perp}=1/G_{n}\) for the CLR normalization), scale models can make this an explicit part of the model building process.
While it is beyond the scope of the present article, we note that there are many ways of building scale models that enhance the interpretability of the parameters and assumptions and a detailed description of these points is describe elsewhere [@nixon2023scale].
In this work, we simply use the parameter \(\gamma\) in the default scale model as a term that represents uncertainty in the CLR assumption and note that larger values of $\gamma$ correspond to more uncertainty in this assumption. In the metatranscriptome example and elsewhere [@Nixon2024B], we also show that the ratio of scales between conditions can be used to build a full scale model when the conditions have dramatically different underlying scales.

>>>>

The analysis of  HTS data suffers from several known problems that can be
traced, in whole or in part, to  misspecification of scale. The first issue 
is poor control of the false discovery rate (FDR) [@Thorsen:2016aa;@Quinn:2018aa;Nearing:2022aa;@Li:2022aa], 
  exhibited as dataset-dependent 
FDR control.  The FDR problem is connected to the double-filtering approach
that is often used, but which is known not to have appropriate FDR control 
[@Zhang:2009aa;@Ebrahimpoor:2021aa]. The second issue is poor performance when 
analyzing asymmetric data; data that arises because of a bias in the direction of change.
This type of data frequently arises in in-vitro selection experiments (SELEX), 
transcriptome analysis, and microbiome analysis [@Wu2021]. The third issue is that these
problems become more pronounced as more samples are collected; that is,
more information results in a worsening of the accuracy of the analysis [@nixon2023scale,@Nixon2024B]. 
The final problem is that these data are relative data, i.e. compositional [@lovell:2011;@Quinn:2019aa], 
and substantial effort has been put into removing this constraint.

@@

Recently, Nixon et al. [-@nixon2023scale] showed that the challenge of non-biological variation in sequencing depth be viewed as a problem of partially-identified models. They showed that \emph{all} normalizations make some assumption about  scale. These implicit assumptions are often  difficult to interpret, and different normalizations provide different outputs when applied to the same dataset [@Bullard:2010;@Dillies:2013;@Thorsen:2016aa;@Weiss:2017aa;@Nearing:2022aa]. Intuitively, normalizations in widespread use  assume that either all samples have the same scale, e.g. proportions, rarefaction [@Hughes:2005tu], RPKM [@Mortazavi:2008; @wagner:tpm], etc; or that a subset of features in one sample can be chosen as a reference to which the others are scaled e.g. the TMM [@Robinson:2010a], or LVHA [@Wu2021] or the additive log-ratio [@aitchison1982]; or that different sub-parts of each sample maintain a constant scale across samples e.g. the RLE [@Anders:2010];  or that the geometric mean of the parts is appropriate e.g the CLR  and its derivatives.

The original ALDEx2 [@fernandes:2013] model unwittingly made a strict assumption about scale through the CLR normalization [@nixon2023scale]. While this assumption could be useful in many cases it could  never be exactly true, and others have shown that it is not always the best option [@Yerke:2024aa]. Nixon et al. [-@nixon2023scale] showed that better scale assumptions resulted in more reproducible data analysis including better control of both false positive and false negative results. In essence ALDEx2 has been modified to explicitly model the scale over a range of reasonable normalization parameters. Here, we briefly introduce these modifications and then show how scale uncertainty can greatly improve modeling in transcriptome and meta-transcriptome datasets to provide more robust and reproducible results.

@@

All  the normalizations in current use are data transformations that can be stated as ratios of the form \(\hat{{\mathbf{W}}}_{dn}=\mathbf{Y}_{dn}/f(\mathbf{Y})\), where the denominator is determined by some function of the observation. We use the \(\hat{{\mathbf{W}}}\) notation to indicate that the normalization is attempting to provide an estimate of the true value.  The technical variation in sequencing depth (\(\mathbf{Y}^{\perp}_{n}=\sum_{d}\mathbf{Y}_{dn}\)) implies that observed data \(\mathbf{Y}\) provides us with information about the system composition \(\mathbf{W}^{\parallel}\) but little to no information in the system scale \(\mathbf{W}^{\perp}\) (Lovell et al. 2011).

@@

The ALDEx2 R package [@fernandes:2013;@fernandes:2014] is a general purpose toolbox for Bayesian modeling of HTS data. For brevity, we discuss ALDEx2 in its simplest form as a tool for estimating the magnitude and statistical significance of Log-Fold-Changes (LFC) (e.g., differential abundance or differential expression analysis), but note that it can be used to fit more complex linear models. At a high-level, ALDEx2 involves three steps: 1) a Bayesian model is used to estimate \(\mathbf{\hat{W}}^{\parallel}\) given observations \(\mathbf{Y}\); 2) a normalization is used to estimate \(\mathbf{\hat{W}}\) given the estimate \(\mathbf{\hat{W}}^{\parallel}\); 3) the \(\mathbf{\hat{W}}\) estimates are used to estimate LFC as part of differential abundance (or expression) analyses. For more details on ALDEx2 see [@fernandes:2013;@nixon2023scale].

By default, ALDEx2 uses the CLR normalization which can be written as \(\log \mathbf{\hat{W}}_{dn}=\log (\mathbf{\hat{W}}^{\parallel}_{dn}/G_{n})\) where \(G_{n}\) denotes the geometric mean of the vector \((\hat{W}^{\parallel}_{1n}, \dots, \hat{W}^{\parallel}_{Dn})\). The CLR normalization equates to an implicit assumption that \(\mathbf{\hat{W}}^{\perp}_{n}=1/G_{n}\)[@nixon2023scale;@Nixon2024B] and even slight errors in this assumption introduces bias into the LFC estimate that is not accounted for when estimating uncertainty (e.g., confidence intervals or credible sets). In fact, Nixon et al. [-@nixon2023scale] showed that the only way in which the ALDEx2 model, or any normalization-based model, could ever be calibrated (e.g., control Type-I Error rates) was if this assumption was exactly true.  In support of this assertion it is known that false discovery rates vary widely by analysis, dataset, and normalization method [@hawinkel2017;@Li:2022aa,Nixon2024B]. To fix this, Nixon et al. [-@nixon2023scale] showed that models should incorporate potential error in the assumptions implied by normalizations.

@@

Many groups have conducted benchmarking studies on different tools and normalizations used for the analysis of datasets such as transcriptomes[@Bullard:2010;@Soneson:2013;@Schurch:2016aa;@Quinn:2018aa;@Li:2022aa] and microbiomes [@McMurdie:2014a;@Thorsen:2016aa; @Weiss:2017aa; @hawinkel2017;@Nearing:2022aa;@Yerke:2024aa]. Generically, it is observed that the actual agreement between analysis methods can be modest, and  no single method outperforms all others in every dataset or type of experiment.  That different tools appear to work more reliably in different datasets from different sources can be explained as different  normalizations being a better fit to the scale of a particular dataset by chance.

In the analysis of HTS data it is often observed that larger datasets converge on the majority of parts being significantly different [@Schurch:2016aa;@Li:2022aa;@nixon2023scale], and that different analytic approaches result in different parts being chosen as significantly different [@Soneson:2013;@Schurch:2016aa;@Quinn:2018aa;@Nearing:2022aa]. Li et al. -@Li:2022aa conducted a permutation-based benchmarking study and found that popular tools performed worse than simple Wilcoxon rank-sum tests in controlling the FDR when sample sizes became large. They suggested that the presence of outliers were one of the factors driving this observation. Brooks et al. -@Brooks:2024aa suggested that inappropriate choice of benchmarking methods are also a major contributing factor. From our perspective, the disagreement between tools can be explained by the partially identifiable nature of the data; that is,  the  observed data are consistent with multiple ways of estimating the parameters [@nixon2023scale]. Partially identifiable data exhibit multiple pathologies, chief among them being that that different analytic approaches can produce  different parameter estimates and that more data produces worse estimates because the additional data increases the precision of a flawed estimate [@gustafson2015bayesian].

Scale simulation  is now build into ALDEx2 to address the issue of partially identifiability, and through this lens the two main root causes to common HTS data pathologies can be proposed. The first contributing factor is the observed very low dispersion estimate for many features that is a by-product of experimental design, sequencing and normalization. In the Schurch et al. @Schurch:2016aa dataset, the data were  from single colonies derived from a single culture. Thus, it is more accurate to describe the 96 samples as wet-lab technical replicates rather than independent samples. However, this type of replication approach is standard in the molecular literature, and would be expected to result in the very low dispersion that is observed.

@@

While the actual scale of the underlying environment is inaccessible post-sequencing, we can study the sensitivity of the results to the choice of normalization and thus scale.
Varying amounts of scale uncertainty is incorporated using the \texttt{gamma} parameter that controls the amount uncertainty added to the CLR mean assumption when we call either \texttt{aldex()}, or \texttt{aldex.clr()}.
The ALDEx2 package  contains a sensitivity analysis function, \texttt{aldex.senAnalysis()}, that can be used to explore the effect of different amounts of scale uncertainty, and an example is shown in  Supplementary Figure 2 for the yeast transcriptome dataset. Here it is clear that even tiny amounts of scale preclude the majority of transcripts from being considered statistically significant.
In practice, we suggest that a \texttt{gamma} parameter between 0.5 and 1 is realistic for most experimental designs, but further work is needed.

@@

Beyond concerns of fidelity and rigor, scale models also enhance the reproducibility and transparency of HTS analyses. The addition of scale uncertainty essentially tests the model over a range of normalizations [@nixon2023scale] and so can replace the consensus approach that has been proposed by some groups [@Nearing:2022aa;@Song:2023aa] with no additional computational overhead.
Thus, an advantage of incorporating scale is that analyses can be made much more robust such that actual or potential differences in scale can be tested and accounted for explicitly.
This occurs because rather than using an implicit assumption (e.g., \(\log \mathbf{W}^{\perp}=1/G_{n}\) for the CLR normalization), scale models can make this an explicit part of the model building process.
While it is beyond the scope of the present article, we note that there are many ways of building scale models that enhance the interpretability of the parameters and assumptions and a detailed description of these points is describe elsewhere [@nixon2023scale].
In this work, we simply use the parameter \(\gamma\) in the default scale model as a term that represents uncertainty in the CLR assumption and note that larger values of $\gamma$ correspond to more uncertainty in this assumption. In the metatranscriptome example and elsewhere [@Nixon2024B], we also show that the ratio of scales between conditions can be used to build a full scale model when the conditions have dramatically different underlying scales.
